Thesis: draft

### 2021-03-26 draft.tex
made some small adjustements in SBM section

{\large \today}\\[4cm] % Date
%\includegraphics[scale=0.7]{ugent_logo.png} % University/department logo - uncomment to place it

\vfill
\end{center}
\end{titlepage}

@@

Another random graph model is one that produces networks containing a community structure, namely the stochastic block model (SBM). The stochastic block model originates in social sciences and was developed to describe group structures in friendship networks \cite{Funke2019}. The idea behind the development of this model is that often in real networks, there are groups of nodes that are more densely interconnected with each other than with the rest of the network \cite{F.Costa2007}. Hence, these networks posses a modular/community structure and this is exactly the structure that is easily modeled by the stochastic block model. The stochastic block model is also often used as a benchmark for community detection algorithms.\\

@@

	\item A partition of the node set $\{1, 2, ..., N\}$ into disjoint subsets $C_1, ..., C_k$; these groups are the communities

@@

There are no real restrictions to the edge probabilities, which makes the stochastic block model a very general model that is capable of reproducing many different network structures. However, the edge probabilities inside a community are often taken to be bigger than the probabilities between communities ($M_{ii} > M_{ij}$, $i \neq j$). In this way, the stochastic block model generates a network with a clear community structure. A stochastic block model that fulfills this condition is called an assortative SBM \cite{Lee2019}. A disassortative stochastic block model refers to the case where the edge probabilities between the communities are higher than those inside the communities ($M_{ii} < M_{ij}$, $i \neq j$) \cite{Gribel2020}.\\

@@

The stochastic block model includes the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. This can be seen in the two different ways. First, consider the case where $k = 1$, thus we only have one group. The matrix $M$ then reduces to one number $p$ and we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. Second, assume we have more than one group or community, $k > 1$, and assume that every element of the matrix $M$ is the same, $M_{ij} = p$ \cite{Clauset2017}. Again, we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model. In both cases, the mathematical properties of the Erd\H{o}s-R\'{e}nyi model are obtained: a Poisson degree distribution, an average distance that is logarithmic with the system size $N$, vanishing clustering coefficient for large $N$, etc \cite{Clauset2017}. The structure within the groups or communities of the stochastic block model is that of a random graph, while the structure between the communities is that of a random bipartite graph \cite{Clauset2017}.\\

@@

One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence of this is that if a network possesses a more skewed degree distribution, the SBM will put the nodes with a high degree in a small group, together with a high probability of connecting to other, larger groups \cite{Clauset2017}. The reason behind this is one of the main assumptions of the stochastic block model, namely the fact that the edge probabilities only depend  on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically  equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. However, a network with a skewed degree distribution violates this assumption of edge independence and the SBM seeks to explain this over other kinds of latent group structures that we might care about \cite{Clauset2017}. Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us too far astray, but interested readers can find more detailed information in reference \cite{Karrer2011}. \\

>>>>

{\large \today}\\[4cm] % Date
%\includegraphics[scale=0.7]{ugent_logo.png} % University/department logo - uncomment to place it
% https://www.overleaf.com/project/605b4c033a439c5f0f15c7e5 --> copywright?
\vfill
\end{center}
\end{titlepage}

@@

The stochastic block model (SBM) is another random graph model that is able to produce networks containing a community structure. The stochastic block model originates in social sciences and was developed to describe group structures in friendship networks \cite{Funke2019}. The idea behind the development of this model is that often in real networks, there are groups of nodes that are more densely interconnected with each other than with the rest of the network \cite{F.Costa2007}. Hence, these networks posses a modular/community structure and this is exactly the structure that is easily modeled by the stochastic block model. The stochastic block model is also often used as a benchmark for community detection algorithms \cite{Abbe2018}.\\

@@

	\item A partition of the node set $\{1, 2, ..., N\}$ into $k$ disjoint subsets $C_1, ..., C_k$; these groups are the communities

@@

There are no real restrictions to the edge probabilities, which makes the stochastic block model a very general model that is capable of reproducing many different network structures \cite{Karrer2011}. However, the edge probabilities inside a community are often taken to be bigger than the probabilities between communities ($M_{ii} > M_{ij}$, $i \neq j$). In this way, the stochastic block model generates a network with a clear community structure. A stochastic block model that fulfills this condition is called an assortative SBM \cite{Lee2019}. A disassortative stochastic block model refers to the case where the edge probabilities between the communities are higher than those inside the communities ($M_{ii} < M_{ij}$, $i \neq j$) \cite{Gribel2020}.\\

@@

The stochastic block model includes the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. This can be seen in the two different ways. First, consider the case where $k = 1$, thus we only have one community and all nodes belong to this community. The matrix $M$ then reduces to one number $p$ and we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. Second, assume we have more than one group or community, $k > 1$, and assume that every element of the matrix $M$ is the same, $M_{ij} = p$ and $M_{ii}=p$ \cite{Clauset2017}. Again, we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model. In both cases, the mathematical properties of the Erd\H{o}s-R\'{e}nyi model are obtained: a Poisson degree distribution, an average distance that is logarithmic with the system size $N$, vanishing clustering coefficient for large $N$, etc \cite{Clauset2017}. The structure within the groups or communities of the stochastic block model is that of a random graph, while the structure between the communities is that of a random bipartite graph \cite{Clauset2017}.\\

@@

One of the shortcomings of fitting the SBM to infer the community structure, is that real-world networks often have a heterogeneous or power-law degree distribution. Since the SBM is made of random networks inside each group and random bipartite networks between each pair of groups, the degree distribution of the full network is always a mixture of Poisson distributions \cite{Clauset2017}. The consequence of this is that if a network possesses a more skewed degree distribution, then fitting the SBM to the data will most likely give poor results \cite{Karrer2011}. The reason behind this is one of the main assumptions of the stochastic block model, namely the fact that the edge probabilities only depend  on which group the node belongs to \cite{Clauset2017}. In other words, nodes that belong to the same group are stochastically  equivalent, meaning that they have equivalent connectivity patterns to other nodes \cite{Clauset2017}. However, a network with a skewed degree distribution violates this assumption of edge independence \cite{Clauset2017}. To say it in the words of Karrer and Newman: "Just as the fitting of a straight line to intrinsically curved data is likely to miss important features of the data, so a fit of the simple stochastic block model to the structure of a complex network is likely to miss much [...]." (Karrer and Newman, p.1, reference \cite{Karrer2011}). Karrer and Newman proposed a variation of the SBM, the degree corrected SBM, that deals with this limitation. We will not discuss their variation, since that would lead us too far astray, but interested readers can find more detailed information in reference \cite{Karrer2011}. \\
